\section{Related Work}
\label{sec:related}

\paragraph{Supervised LLM routing.}
RouteLLM learns routers from human preference data and augmented pairwise labels~\citep{ong2024routellm}.
HybridLLM trains a query encoder with supervision from model quality scores~\citep{ding2024hybrid}.
GraphRouter routes among models and agents with graph-based supervised training~\citep{feng2025graphrouter}.
AgentRouter formulates multi-agent QA as knowledge-graph--guided routing, training a heterogeneous GNN with soft supervision from per-agent performance~\citep{zhang2026agentrouter}.
RouterBench provides benchmarks for comparing supervised routers~\citep{hu2024routerbench}.
These systems \emph{learn} routing inputs from preference data, outcome labels, or structured supervision; our contrast is at \textbf{signal extraction}: we pre-specify unsupervised statistics and ask whether they predict oracle $r(q)$ (appropriate model choice) before any routing-specific representation learning.

\paragraph{Routing vs.\ cascading (related work).}
Surveys distinguish \textbf{routing}---choose one model per query---from \textbf{cascading}---sequentially invoke models with feedback~\citep{chen2024frugalgpt}.
FrugalGPT and related work use cascades or scorers after partial generations; we study \textbf{binary model selection} from pre-specified signals, not a new cascade algorithm.
Smoothie combines multiple model outputs with label-free weighting over candidates~\citep{guha2024smoothie}.
Such methods often require post-generation information, multiple samples, or reward models; we focus on \textbf{pre-decision signals} from query text and model responses in a fixed capability-differentiated pool, and we report accuracy--cost \textbf{Pareto} curves rather than accuracy alone.

\paragraph{Uncertainty and confidence signals.}
Work on calibration and self-evaluation shows that LLM probabilities can relate to correctness~\citep{kadavath2022know,plaut2024calibration,he2023calibration}.
Semantic uncertainty uses multiple samples to estimate confidence~\citep{kuhn2023semantic}.
These lines motivate \textbf{model-dependent} model-response and \textbf{cross-model} comparative signals but do not systematically test model-independent, model-dependent, and cross-model layers \textbf{in parallel} against oracle $r(q)$---$\Mlo$ fails, $\Mhi$ succeeds---nor separate signal informativeness (H1--H3) from deployable \textbf{binary routing} quality (H4).

\paragraph{Positioning and survey taxonomy.}
Recent routing surveys distinguish systems by \textbf{when} routing occurs (before vs.\ during vs.\ after generation), \textbf{what information} the router may use (query only, single-model outputs, multiple models), and \textbf{how} the routing score is formed (rules, uncertainty, learned models).
Table~\ref{tab:contrast} summarizes our contrast with supervised routers; our signal layers (Section~\ref{sec:signal-extraction}) align with the information axis: model-independent $\phi$ (query-side), model-dependent $\psi$ (weak-model-side), cross-model $\chi$ (multi-model-side).
We separate the routing \textbf{objective} (oracle $r(q)$: was $\Mhi$ appropriate?) from the routing \textbf{mechanism} ($\pi(q)$: select pool member from $x(q)$) and evaluate under accuracy--cost Pareto curves.

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{@{}p{0.42\linewidth}p{0.48\linewidth}@{}}
    \toprule
    \textbf{Typical supervised router} & \textbf{This work} \\
    \midrule
    Router input: classifier on query ($+$ history), trained on prefs or outcomes &
    \textbf{Unsupervised signals} $\rightarrow$ feature vector $x(q)$ \\
    Labels \textbf{define} the router end-to-end &
    Labels used \textbf{only on calib}: signal analysis + $(\lambda,\tau)$ fit \\
    Signal source: learned representations from routing data &
    Query-derived (model-independent), model-response (model-dependent), cross-model comparative (Section~\ref{sec:signal-extraction}) \\
    Evaluation: routing accuracy or win rate &
    \textbf{Pareto} over task $\mathrm{Acc}(\pi)$ and $\mathrm{Cost}(\pi)$ on test \\
    \bottomrule
  \end{tabular}
  \caption{Contrast with supervised LLM routing (program §2). Unsupervised describes layer~1 signal extraction; calib uses offline $r(q)$ for analysis and policy fit only.}
  \label{tab:contrast}
\end{table}
