\subsection{Signal extraction}
\label{sec:signal-extraction}

We inventory \textbf{unsupervised} candidate features---no $r(q)$ at extraction time---organized into three \textbf{signal layers} (Section~\ref{sec:intro}).
Each layer rests on a \textbf{signal assumption} (why it might predict oracle $r(q)$); H1--H3 test whether each layer helps \textbf{binary routing} on its own.

\paragraph{Signal layers.}
\begin{enumerate}
  \item \textbf{Model-independent} --- \emph{query-derived} signals (H1).
  \item \textbf{Model-dependent} --- \emph{model-response} signals (H2).
  \item \textbf{Cross-model} --- \emph{cross-model comparative} signals (H3).
\end{enumerate}

\paragraph{Signal assumptions.}
\begin{itemize}
  \item \textbf{Query-derived (model-independent):} More complex or ambiguous queries are more likely to require a higher-capability pool member $\Rightarrow$ \textbf{H1}.
  \item \textbf{Model-response (model-dependent):} Higher model uncertainty reflects higher probability of error on that query $\Rightarrow$ \textbf{H2}.
  \item \textbf{Cross-model comparative (cross-model):} Differences between pool members on the same query reveal routing opportunities $\Rightarrow$ \textbf{H3}.
\end{itemize}
Assumptions are priors, not claims; a rejected hypothesis means the assumption failed for our locked setting.

\paragraph{Query-derived (model-independent).}
Computed from query text only---no forward pass through any \textbf{routing-pool} member at extraction (a fixed sentence encoder for $u(q)$ is allowed; it is not a member of $\mathcal{P}$): e.g., lexical/syntactic complexity, length, ambiguity cues.
Semantic representation $u(q)$ is stored separately; scalar $\phi(q)$ includes corpus-relative \emph{embedding-geometry} features derived from $u(q)$ (fit on $R_c$ only).
Token counts use the canonical user prompt the pool model receives.
\texttt{compression\_ratio} measures text \emph{redundancy/compressibility} (zlib), not semantic complexity.
PCA coordinates (\texttt{pc1--pc3}) describe position on the calibration embedding manifold---not an interpretable difficulty axis.

\paragraph{Model-response (model-dependent).}
Computed from one pool member on $q$: e.g., token entropy $H(q\mid M_i)$, max softmax probability (MSP), margin, log-probability.
On our locked MCQ protocol, $\psi$ uses \textbf{choice-letter logits} at the answer token (entropy, MSP, margin over $\{A,B,C,D\}$); open-ended protocols would use different extractors (sequence log-probability, sample-based uncertainty) under the same layer.

\textbf{MSP is a routing signal, not a calibrated probability.}
We store MSP in \texttt{prediction.confidence} for convenience, but \textbf{H2} asks whether the model's \emph{confidence signal} predicts routing need $r(q)$---not whether MSP is an accurate estimate of $P(\text{correct})$.
Chat and instruction-tuned LMs are often overconfident; softmax masses are useful as \emph{ranking} features for routing analysis even when miscalibrated~\citep{plaut2024calibration}.

\paragraph{Cross-model comparative (cross-model).}
Computed by comparing pool members on the same $q$, typically from model-response quantities already computed: e.g., entropy gap $\Delta H$, confidence gap, answer disagreement.
\textbf{H3} tests $\chi(q)$ \emph{alone}---not whether $\phi{+}\psi$ beats $\phi$ (combination analysis belongs in signal selection, Section~\ref{sec:signal-selection}).

Table~\ref{tab:setup} lists concrete signals once the setting is locked (Stages 1--3).
\textbf{Layers are universal;} concrete features are \textbf{protocol-dependent implementations} for the locked Experimental Setting $\mathcal{S}$.

The output of this stage is a raw signal inventory per query---not yet selected into $x(q)$.
