\subsection{Signal analysis}
\label{sec:signal-analysis}

\textbf{Purpose.} Test whether each \emph{signal representation} carries predictive information about oracle $r(q)$ (appropriate model choice)---independent of routing policy quality.
Stage~6 **measures informativeness** only; it does **not** select features (Stage~7) or train a routing policy (Stage~8).

\textbf{Split.} Calibration set $R_c$ only; labels are offline $r(q)$ from Equation~\eqref{eq:routing-need}.

\paragraph{Artifacts (Stage~6).}
Unified \texttt{analysis\_table.csv} on $R_c$ (hierarchical \texttt{phi.*}, \texttt{psi.*}, \texttt{chi.*} columns); primary \texttt{representation\_tests.json}; diagnostic \texttt{univariate\_\{query,response,cross\}.jsonl}; manifest \texttt{signal\_analysis\_meta.json} (schema version, seed, dataset, full artifact list).

\subsubsection{Representation evaluation (6.1)}
\label{sec:signal-analysis-representation}

\textbf{Primary.} For each pre-specified representation, fit an evaluation model (default: 5-fold stratified CV logistic, \texttt{class\_weight=balanced}) vs.\ $r(q)$ on $R_c$.
Report mean CV AUROC / AUPRC and per-fold scores (Table~\ref{tab:signal-analysis}; artifact \texttt{representation\_tests.json}).

\begin{itemize}
  \item \textbf{H1a} --- structural $\phi_{\text{structural}}$
  \item \textbf{H1b} --- ambiguity $\phi_{\text{ambiguity}}$
  \item \textbf{H1c} --- embedding geometry $\phi_{\text{geometry}}$
  \item \textbf{H1} --- combined model-independent $\phi(q)$
  \item \textbf{H2} --- model-response $\psi(M_{\mathrm{lo}})$
  \item \textbf{H3} --- cross-model comparative $\chi$ (continuous metrics in joint CV)
\end{itemize}

\textbf{H1 wording:} evaluate each $\phi$ \emph{signal family} individually, then the full $\phi$ representation---not single-scalar hypotheses.

\subsubsection{Feature diagnostics (6.2)}
\label{sec:signal-analysis-diagnostics}

\textbf{Secondary.} Per-scalar Spearman $\rho$, AUROC, AUPRC, and class-conditional means on $R_c$ (\texttt{univariate\_\{query,response,cross\}.jsonl}).
These support interpretation and feed Stage~7; they are \textbf{not} the primary H1--H3 tests.

\paragraph{Hypotheses (motivated by signal assumptions, Section~\ref{sec:signal-extraction}).}
\textbf{H1 (model-independent):} Query-derived representations predict $r(q)$.
\textbf{H2 (model-dependent):} Model-response signals predict $r(q)$---$\psi(q,\Mlo)$ \emph{alone}.
\textbf{H3 (cross-model):} Cross-model comparative signals predict $r(q)$---$\chi(q)$ \emph{alone}.

Each hypothesis is tested on \emph{its own layer's representation}, not as incremental gain over the previous layer and not as $\phi{+}\psi$ combination (that supports Stage~7).

We do \textbf{not} report $\pi(q)$ or accuracy--cost at this stage.
Results appear in Section~\ref{sec:results-signals}.
