\subsection{Signal selection}
\label{sec:signal-selection}

After Stage~6 signal analysis on calib, Stage~7 freezes a compact feature vector $x(q)\in\mathbb{R}^d$ for all routing experiments.
This stage corresponds to \textbf{Section~7} in the pipeline narrative: ranking, redundancy, correlation, and export---not hypothesis testing.

\subsubsection{Feature ranking (7.1)}
\label{sec:signal-selection-ranking}

Rank per-scalar diagnostics from Stage~6 univariates (e.g.\ $|$\texttt{auroc}$-0.5|$ on $R_c$).
Ranking is \textbf{not} performed in Stage~6 outputs.

\subsubsection{Redundancy and correlation (7.2)}
\label{sec:signal-selection-redundancy}

Identify highly redundant pairs (Pearson $|r|\ge 0.95$ on $R_c$) within $\phi$ blocks and across candidates.
Correlation matrices support drop decisions; they are Stage~7 diagnostics, not Stage~6 representation tests.

\subsubsection{Freeze $x(q)$ (7.3)}
\label{sec:signal-selection-freeze}

\paragraph{Rules.}
\begin{itemize}
  \item \textbf{Do not pre-prune} before Stage~6 univariate results exist.
  \item Drop features with evidence only: AUROC $\approx 0.5$, Spearman $\approx 0$, zero variance, or redundancy ($|r|>0.95$ on $R_c$).
  \item Drop redundant features (high correlation, no $\Delta$AUROC gain).
  \item Drop non-informative features unless needed for the H1--H3 ablation narrative.
  \item \texttt{option\_count} is not extracted (constant on ARC MCQ).
  \item Export a frozen feature specification used identically on calib and test (values only; no refitting).
\end{itemize}

Selection uses calib analysis only---not test routing performance.
\input{tables/T3_feature_selection}
