\subsection{Signal analysis (H1--H3)}
\label{sec:results-signals}

We report whether each \emph{signal representation} predicts oracle $r(q)$ on calib before any \textbf{binary routing} policy claims.
H1--H3 are \textbf{parallel representation tests}: each asks whether that layer's feature vector helps predict $r(q)$, not whether one layer beats another and not whether $\phi{+}\psi$ combination defines H3 (H3 = $\chi$ alone; combinations support Stage~7 selection).

\subsubsection{Representation evaluation (6.1)}
\label{sec:results-signals-representation}

For each representation on $R_c$, we fit a logistic evaluation model (5-fold stratified CV) and report mean CV AUROC and AUPRC vs.\ $r(q)$, with per-fold scores stored in \texttt{representation\_tests.json} for stability analysis.
Representations: \textbf{H1a} structural, \textbf{H1b} ambiguity, \textbf{H1c} embedding geometry, \textbf{H1} combined $\phi$, \textbf{H2} $\psi(M_{\mathrm{lo}})$, \textbf{H3} $\chi$.
Table~\ref{tab:signal-analysis} is the primary evidence for H1--H3.
\textsc{[Values from \texttt{representation\_tests.json}.]}

\subsubsection{Feature diagnostics (6.2)}
\label{sec:results-signals-diagnostics}

Per-feature Spearman $\rho$, AUROC, AUPRC, and class-conditional means ($r{=}0$ vs.\ $r{=}1$) from \texttt{univariate\_\{query,response,cross\}.jsonl} support interpretation of \emph{why} a representation may work; they are not the primary hypothesis tests.
\textsc{[Per-feature appendix tables TBD.]}

\input{tables/T2_signal_analysis}

\paragraph{Summary.}
Table~\ref{tab:signal-analysis} asks whether each representation---not each individual scalar---contains routing information.
A positive H2 with a weak H1 is valid---model-dependent uncertainty may carry routing information without query-side cues.
Feature ranking, redundancy pruning, and frozen $x(q)$ are Stage~7 (Section~\ref{sec:signal-selection}).
