% §1 Introduction

\section{Introduction}
\label{sec:intro}

Large language models (LLMs) vary widely in capability and inference cost, and no single model optimally balances response quality and computational expense across all queries \citep{hu2024routerbench,ong2024routellm}.
Different queries require different levels of reasoning capability; routing every query through the same model is therefore inefficient \citep{ding2024hybrid,chen2023frugalgpt}.
Engineers deploy multi-LLM systems that must trade off quality against cost when selecting among models in a fixed pool $\mathcal{M}$ \citep{hu2024routerbench,ding2024hybrid}.

\paragraph{Problem.}
Query routing assigns each input $q$ to a model in $\mathcal{M}$.
Prior work largely follows one of three patterns: \emph{supervised routers} trained on preference or routing labels \citep{ong2024routellm,ding2024hybrid,feng2025graphrouter}; \emph{learned pre-hoc estimators} of model quality \citep{cao2026scope}; or \emph{post-generation} selection after every pool member has produced a full answer \citep{guha2024smoothie,niu2026cascal}.
These systems optimize \emph{which model to run}; they do not systematically characterize whether inexpensive \textbf{unsupervised pre-inference signals}---computable from the query and from a lightweight prefill probe, without routing supervision labels---themselves carry routing-relevant information before expensive full generation.

\paragraph{Our approach.}
We study \textbf{routing based on unsupervised pre-inference signals}.
For each query we extract a feature vector from two complementary families: model-independent statistics $c(q)$ from formatted query text alone, and model-dependent probe statistics $H(q,M_i)$ and $m(q,M_i)$ from one prefill forward pass per pool member (\S\ref{sec:method}).
We then ask how much \textbf{routing-relevant information} these signals contain, whether the families complement each other, and---only after that characterization---whether a simple calibrated policy can exploit whatever information exists.
Our primary contribution is \textbf{signal extraction and characterization}, not a deployable router; Study~IV is a \textbf{demonstration} built on the signals Studies~I--III establish.

\paragraph{Contrast with prior routing pipelines.}
\begin{center}
\small
\begin{tabular}{@{}l@{\quad}l@{}}
\textbf{Typical supervised routing} &
Query $\rightarrow$ supervised router $\rightarrow$ choose LLM \\
\textbf{This work} &
Query $\rightarrow$ unsupervised signal extraction $\rightarrow$ characterize information $\rightarrow$ simple policy $\rightarrow$ choose LLM \\
\end{tabular}
\end{center}

\paragraph{Research question.}
\textit{How informative are unsupervised pre-inference routing signals for multi-LLM selection before full model generation?}

Scientifically, we ask how much \textbf{routing-relevant information} is present in inexpensive pre-inference signals.
We operationalize \emph{informativeness} as predictive association with routing need (Spearman $\rho$, opportunity AUROC, complementary predictive gain $\Delta$AUROC)---not as sufficiency for optimal routing or as Shannon mutual information.
Weak or uniformly null associations are valid findings if reported rigorously \citep{plaut2024chat}; in the latter case the paper emphasizes \emph{limits} of these signal families under our protocol rather than enabling routing (\S\ref{sec:discussion}).

\paragraph{Appropriate selection.}
Within $\mathcal{M}$, appropriate assignment means a cost--quality trade-off: route to the \textbf{weakest model expected to answer correctly}, escalating to a stronger model only when pre-inference signals indicate that additional capability is likely to improve the outcome---not always choosing the highest-capability model.

\paragraph{Scientific pipeline.}
Our contribution follows an \textbf{information $\rightarrow$ characterization $\rightarrow$ understanding $\rightarrow$ utility} structure:
\begin{enumerate}
  \item \textbf{Signal taxonomy} --- two families in $\mathcal{S}=\mathcal{S}_{\text{indep}}\cup\mathcal{S}_{\text{dep}}$; v1 instances $c(q)$, $H$, $m$ (\S\ref{sec:method}).
  \item \textbf{Characterization} (Studies I--II; RH1--RH2) --- does each family encode routing-relevant information?
  \item \textbf{Understanding} (Study III; RH3) --- do the families provide \emph{additional} information together?
  \item \textbf{Utility} (Study IV; RH4) --- can a simple policy \emph{exploit whatever information exists} under a cost--quality trade-off?
\end{enumerate}
Each study depends on the previous; routing evaluation is meaningful only if Studies~I--III establish whether exploitable information is present.

\paragraph{Supervision boundary.}
Signal \textbf{computation} is unsupervised: $c(q)$, $H$, and $m$ are extracted without routing labels.
\textbf{Characterization} uses offline oracle labels from full generation on held-out data only to measure informativeness---never to define signals \citep{hu2024routerbench}.
\textbf{Policy calibration} (Study~IV) fits a simple logistic model and threshold on the validation split from oracle-derived $y^{\text{opp}}$; this supervised step is a secondary demonstration on top of unsupervised signals, not an end-to-end supervised router (\S\ref{sec:method:supervision}).

\paragraph{Gap.}
Table~\ref{tab:landscape} (\S\ref{sec:related}) summarizes prior clusters.
None systematically characterize whether model-independent and model-dependent \textbf{pre-inference signal families} jointly encode routing need at prefill cost rather than full generation cost \citep{guha2024smoothie}.

\paragraph{Contributions.}
In summary:
\begin{enumerate}
  \item We formalize \textbf{unsupervised pre-inference signal extraction and characterization} for multi-LLM routing, with a three-layer supervision boundary between signal computation, offline evaluation, and policy demonstration (\S\ref{sec:method}).
  \item We define model-independent and model-dependent signal families, a shared prompt--probe--oracle protocol, and operational informativeness metrics (\S\ref{sec:method}--\S\ref{sec:setup}).
  \item We characterize per-family routing-relevant information on $n{=}1{,}172$ ARC-Challenge test queries (Studies I--II; \S\ref{sec:characterization}).
  \item We evaluate complementary predictive information across families (Study~III; \S\ref{sec:char:study3}).
  \item We test whether a simple calibrated policy can exploit the information characterized in Studies~I--III (Study~IV; \S\ref{sec:routing}).
\end{enumerate}

\paragraph{Main findings.}
\textbf{[TBD: branch on outcome scenario---(a) some signal informative and utility may improve, or (b) uniformly null signals $\Rightarrow$ limits characterization.]}

\paragraph{Paper organization.}
\S\ref{sec:related} reviews related work.
\S\ref{sec:method}--\S\ref{sec:setup} define signals, supervision layers, and protocol.
\S\ref{sec:characterization} reports Studies~I--III.
\S\ref{sec:routing} reports Study~IV.
\S\ref{sec:discussion}--\S\ref{sec:conclusion} conclude.
