% §2 Related Work

\section{Related Work}
\label{sec:related}

Prior query-routing research has focused on learning deployable routers---supervised predictors or unsupervised post-generation selectors---rather than on characterizing what unsupervised information exists before full model generation.
We organize related work by what information is used for routing and when it becomes available, following the cost--quality framing of \citet{hu2024routerbench} and the timing distinction emphasized by \citet{ong2024routellm}.
Table~\ref{tab:landscape} summarizes the landscape; paragraphs below develop each cluster and position our measurement framework.

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{p{0.20\linewidth}p{0.26\linewidth}p{0.18\linewidth}p{0.26\linewidth}}
    \toprule
    \textbf{Category} & \textbf{Routing information} & \textbf{Timing} & \textbf{Representative work} \\
    \midrule
    Supervised routers &
      Annotated preferences, labels, histories &
      Query / post-gen &
      RouteLLM, Hybrid LLM, GraphRouter, FrugalGPT \\
    Unsupervised post-gen &
      Full pool responses &
      After $m$ generations &
      Smoothie, CASCAL, Mu et al. \\
    Pre-hoc estimation &
      Anchor fingerprints; predicted correctness/cost &
      Before target inference &
      SCOPE \\
    Unsupervised pre-gen (ours) &
      Query text; prefill logprobs &
      Before full generation &
      This work \\
    Router evaluation &
      Precomputed outcomes &
      --- &
      RouterBench \\
    \bottomrule
  \end{tabular}
  \caption{Related work organized by routing information and timing.}
  \label{tab:landscape}
\end{table}

\paragraph{Query routing with supervised annotation.}
RouteLLM learns $P(\text{strong wins} \mid q)$ from human preference data and augmented labels, routing at inference from query text without live prefill probes on each candidate LLM \citep{ong2024routellm}.
The system reduces cost between model classes but requires supervised preference construction and does not characterize unsupervised per-model logprob statistics.

Hybrid LLM trains a DeBERTa encoder on when a small model matches large-model quality, using BARTScore on offline generations as supervision \citep{ding2024hybrid}.
Routing reads the query string only; the authors note query-only routing as a limitation and call for richer pre-query signals---motivating our study, though their method remains supervised on quality proxies.

GraphRouter learns over query--LLM edges with PLM embeddings and historical performance and cost \citep{feng2025graphrouter}.
Inference relies on past interaction data, not unsupervised live probes on every candidate.

FrugalGPT optimizes API cascades using a scorer $g(q,\text{answer})$ after each generation step \citep{chen2023frugalgpt}.
Our primary signals are computed before expensive full inference; calibrated margin cascades \citep{kotte2026ucci} provide a complementary production contrast.

RouterBench precomputes outcomes for eleven LLMs across eight datasets and compares router families \citep{hu2024routerbench}.
It formalizes oracle and zero routers and shows that simple predictive routers do not consistently beat the zero router on all tasks---setting a high bar for probe-based policies.
RouterBench does not study unsupervised prefill logprob probes as routing inputs.

\paragraph{Unsupervised query routing without manual annotation.}
Smoothie routes an unlabeled test set by estimating per-model generation quality from post-generation output embeddings via a weak-supervision graphical model \citep{guha2024smoothie}.
In contrast to our setting, Smoothie requires $m$ full generations per query; we study $m$ prefill probes and cost--quality selection within a tiered pool.

CASCAL trains routers without in-domain ground truth using synthetic queries and confidence-weighted consensus across pool responses---again post-generation \citep{niu2026cascal}.

Mu et al.\ propose unsupervised query routing for retrieval-augmented generation: they construct a multi-source upper-bound response to evaluate single-source retrieval quality and thereby assign queries to search engines without manual annotation \citep{mu2025ragrouting}.
Their framework eliminates labeling cost and improves scalability and out-of-distribution generalization, but requires generated responses from every retriever rather than prefill logprobs from every LLM.
Our work asks what can be inferred before any pool member completes an answer.

\paragraph{Pre-hoc performance estimation.}
SCOPE trains a reasoning estimator on anchor-set correctness and cost to predict outcomes before target-query inference \citep{cao2026scope}.
It addresses a \emph{routing-system} question---which model to assign under a budget---whereas we address a \emph{characterization} question---whether lightweight unsupervised signals in $\mathcal{S}$ encode routing need without such an estimator.
We cite SCOPE for shared pre-hoc timing and utility-based evaluation reporting, not as a methodological template.

\paragraph{Partial pre-generation routing.}
UniRoute clusters prompts unsupervisedly but uses supervised per-cluster validation errors per model for routing among changing pools \citep{jitkrittum2025uniroute}.
New models require labeled validation runs and are not fully label-free at deployment.

CP-Router routes between one LLM and one large reasoning model using conformal-prediction set size from MCQ logits \citep{su2025cprouter}---closest in timing to our setting.
It studies a binary LLM$\leftrightarrow$LRM choice, not entropy and margin across a multi-LLM pool under a unified unsupervised protocol.

\paragraph{Uncertainty, calibration, and correctness prediction.}
Kadavath et al.\ study whether LMs know what they know, including MCQ self-evaluation from pre-answer logits \citep{kadavath2022language}.
He et al.\ show instruction-tuned chat models become overconfident on MCQ-style prompts, conflating answer and format uncertainty \citep{he2023uncertainty}.
Plaut et al.\ find chat LLM probabilities miscalibrated on ARC and MMLU yet predict correctness (AUROC); margin and entropy appear in supplementary analysis \citep{plaut2024chat}.
Kuhn et al.\ use multi-sample semantic entropy \citep{kuhn2023semantic}---requiring multi-generation budgets, unlike our single-pass prefill constraint.
Kotte \citep{kotte2026ucci} shows calibrated token-margin cascades outperform raw entropy thresholding in production---a contrast to our uncalibrated prefill probes.
Lugoloobi et al.\ route with supervised linear probes on pre-generation hidden states \citep{lugoloobi2026failures}; timing is similar but the signal class differs from unsupervised logprob statistics in $\mathcal{S}$.

\paragraph{Positioning.}
We do not propose another supervised router, a pre-hoc reasoning estimator \citep{cao2026scope}, nor an unsupervised method that requires $m$ full generations per query before selection \citep{guha2024smoothie,niu2026cascal,mu2025ragrouting}.
Given a fixed pool and frozen prompt protocol, we characterize how informative pre-hoc unsupervised routing signals are---organized by the model-independent and model-dependent signal families---not whether a learned estimator can replace them \citep{cao2026scope}.
Weak or null predictive association is a valid finding \citep{plaut2024chat,hu2024routerbench}.
Study IV (\S\ref{sec:routing}) evaluates decision utility only after Studies I--III establish predictive information.
