% §3 Method

\section{Method}
\label{sec:method}

This section formalizes the measurement framework introduced in \S\ref{sec:intro}.
We define unsupervised pre-inference routing signals, separate three supervision layers (signal computation, characterization, policy demonstration), and specify how we operationalize informativeness.
Experimental configuration, splits, and study hierarchy are in \S\ref{sec:setup}; empirical results follow in \S\ref{sec:characterization}--\S\ref{sec:routing}.

\subsection{Problem formulation}
\label{sec:method:formulation}

Let $\mathcal{M}=\{M_1,\ldots,M_k\}$ be a fixed model pool of instruction-tuned LLMs with known relative cost tiers \citep{hu2024routerbench,ding2024hybrid}.
For each query $q$, routing assigns one $M_i \in \mathcal{M}$.
We study \textbf{unsupervised pre-inference routing signals}: statistics computable before full generation on the target query, without using routing supervision labels to extract them \citep{guha2024smoothie,mu2025ragrouting}.
Throughout, \emph{pre-inference} means the signal is available before paying the inference cost of a complete answer; a \textbf{prefill probe} (one forward pass on the formatted prompt) is permitted for the model-dependent family, but not autoregressive decoding to completion.

\paragraph{Appropriate assignment.}
Appropriate assignment means achieving a cost--quality trade-off within $\mathcal{M}$: route to the \textbf{weakest model expected to answer correctly}, escalating to a stronger model only when pre-inference signals indicate that additional capability is likely to improve the outcome.
Our empirical question is \textbf{how much routing-relevant information} inexpensive pre-inference signals carry for that assignment---not whether a particular router architecture maximizes accuracy.

\paragraph{Research hypotheses.}
Studies I--IV test a sequential chain; Study~IV is conditional on what Studies~I--III establish:
\begin{align*}
  \text{RH1:}& \;\mathcal{S}_{\text{indep}} \text{ contains measurable routing-relevant information} \\
  \text{RH2:}& \;\mathcal{S}_{\text{dep}} \text{ contains measurable routing-relevant information} \\
  \text{RH3:}& \;\text{the families provide additional information beyond either alone} \\
  \text{RH4:}& \;\text{a simple policy can exploit whatever information exists for cost--quality gains.}
\end{align*}

\paragraph{Oracle upper bound (offline only).}
The oracle is an evaluation construct, not a routing method.
For each $(q,M_i)$, we run full inference under a frozen prompt protocol and assign correctness $y(q,M_i)\in\{0,1\}$ using a task-appropriate metric (\S\ref{sec:setup}).
Prefill signals are never used to compute oracle labels.
Among models tied at maximum quality, the oracle prefers the lowest-cost pool member \citep{hu2024routerbench}.
This offline upper bound defines routing need for evaluation only; we do not use it to train a neural router.

\paragraph{Routing need.}
For a designated weak--strong pair $(M_w,M_s)$, query $q$ exhibits \textbf{routing opportunity} when $y(q,M_w)=0$ and $y(q,M_s)=1$.
We write the binary indicator $y^{\text{opp}}(q)\in\{0,1\}$.
Other $(y_w,y_s)$ combinations define \emph{easy} (both succeed), \emph{weak-only} (weak succeeds, strong fails), and \emph{too-hard} (both fail).
These four buckets structure characterization in \S\ref{sec:characterization}.

\subsection{Three layers of supervision}
\label{sec:method:supervision}

We distinguish three stages; only the first is unsupervised in the machine-learning sense:
\begin{table}[t]
  \centering
  \small
  \begin{tabular}{@{}p{0.22\linewidth}cp{0.52\linewidth}@{}}
    \toprule
    \textbf{Stage} & \textbf{Labels?} & \textbf{Role} \\
    \midrule
    Signal computation ($c$, $H$, $m$) & No &
    Unsupervised extraction; \textbf{primary contribution} \\
    Characterization (Studies I--III) & Oracle offline only &
    Measure $\rho$, AUROC, $\Delta$AUROC; never define signals \\
    Policy demonstration (Study IV) & Yes on validation &
    Logistic $+$ threshold from $y^{\text{opp}}$; \textbf{secondary} \\
    \bottomrule
  \end{tabular}
  \caption{Supervision boundary: unsupervised signal extraction vs.\ offline evaluation vs.\ policy calibration.}
  \label{tab:supervision}
\end{table}

\noindent
\textbf{Signal computation} produces $c(q)$, $H(q,M_i)$, and $m(q,M_i)$ without routing labels, preference data, or router training.
\textbf{Characterization} uses oracle labels generated by offline full inference solely to quantify predictive association with routing need.
\textbf{Policy demonstration} fits a simple logistic regression $\Pr(y^{\text{opp}}\mid c, H_w, m_w)$ and tunes a decision threshold $\tau$ (and optional cost penalty $\lambda$) on the ARC validation split; reported utility is evaluated on the held-out test split (\S\ref{sec:routing}).
We do not claim an unsupervised routing \emph{system}: unsupervised describes \emph{signal extraction}; calibration is an explicit, secondary step built on those signals.

\subsection{Signal-family taxonomy}
\label{sec:method:signals}

We group candidate statistics into two \textbf{signal families} forming a pre-inference feature vector:
\begin{equation}
\mathcal{S} = \mathcal{S}_{\text{indep}} \cup \mathcal{S}_{\text{dep}}, \quad
\mathcal{S}_{\text{indep}} = \{ s(q) \}, \quad
\mathcal{S}_{\text{dep}} = \{ s(q, M_i) \}.
\end{equation}
The \textbf{model-independent family} $\mathcal{S}_{\text{indep}}$ contains $s(q)$ computed from formatted query text only (no LLM forward pass).
The \textbf{model-dependent family} $\mathcal{S}_{\text{dep}}$ contains $s(q,M_i)$ from a prefill probe on each pool member.
The taxonomy organizes what we measure; the scientific object is whether these families encode routing-relevant information---including null results.

\paragraph{Model-independent family: representative $c(q)$.}
We study one representative complexity signal $c(q)$, selected by one-time calibration on ARC validation (\S\ref{sec:setup:calib}).
Candidates span four families: token length (\texttt{piece\_count}), lexical diversity (MATTR), lexical information (Shannon entropy over tokenizer pieces), and compressibility (zlib ratio on UTF-8 text).
The winner becomes the frozen $c(q)$ for all test reporting; we do not re-select on the held-out test set or on MMLU.

Lexical Shannon entropy $H_{\text{text}}(q)$ operates on tokenizer piece frequencies---a different random variable from model-dependent probe entropy $H(q,M_i)$ (\S\ref{sec:method:prefill}).

\paragraph{Model-dependent family: $H$ and $m$.}
\label{sec:method:prefill}
We implement token entropy and log-probability margin from one prefill forward pass per $(q,M_i)$:
\begin{equation}
H(q,M_i) = -\sum_{v \in V} p_T(v)\log p_T(v), \qquad
m(q,M_i) = p_T^{(1)} - p_T^{(2)},
\end{equation}
where $p_T$ is the full-vocabulary softmax at probe step $t=T$ (last prompt token), temperature $\tau=1$, and top-$k$/top-$p$ are disabled.
We refer to $m$ as log-probability margin; we do not call $m$ or peak probability \emph{confidence}, which implies calibration \citep{plaut2024chat,kotte2026ucci}.
Auxiliary quantities logged on the same pass include $p_{\max}$ and vocabulary size for auditability.

Raw $H(q,M_i)$ and $m(q,M_i)$ are not assumed comparable across $M_i$ \citep{he2023uncertainty,plaut2024chat}.
Analysis uses within-query comparisons, ranks, and correlation with offline routing need.

\subsection{Operational definition of informativeness}
\label{sec:method:informativeness}

We use informativeness operationally, not as Shannon mutual information $I(S;Y)$.
A signal is informative to the extent it shows \textbf{predictive association with routing need}, measured by:
\begin{itemize}
  \item Spearman $\rho$ with bootstrap CI vs.\ $y^{\text{opp}}$ (and vs.\ oracle gap as secondary);
  \item AUROC / AUPRC for opportunity detection;
  \item complementary predictive gain: $\Delta$AUROC when adding a signal family beyond another (Study~III).
\end{itemize}
If every studied signal yields AUROC $\approx 0.50$ on the corrected test split, we report a \textbf{limits} characterization: what these probe families do \emph{not} provide under our protocol, rather than a practical routing method (\S\ref{sec:discussion}).

\subsection{Prompt protocol}
\label{sec:method:protocol}

Prompt construction has three layers; the same chat prompt $P(u)$ is used for signal extraction and oracle evaluation on a given $(q,M_i)$:
\begin{center}
\small
Benchmark item $\rightarrow$ task formatting ($u$) $\rightarrow$ chat template $P(u)$\\
$\hookrightarrow$ prefill probe (signals) \quad|\quad full inference (oracle labels, offline)
\end{center}

\begin{enumerate}
  \item \textbf{Task formatting} --- Deterministic \texttt{user\_content} per benchmark (e.g., ARC-Challenge: question, labeled choices, letter-only reply instruction).
  \item \textbf{Chat template} --- \texttt{tokenizer.apply\_chat\_template} wraps $u$ in the model-native template without a researcher-defined system prompt.
  \item \textbf{Branch} --- Prefill probe for signals; greedy \texttt{generate()} for oracle labels only.
\end{enumerate}
We log \texttt{protocol\_version}, \texttt{prompt\_hash}, \texttt{prompt\_tokens}, and \texttt{extraction\_method} per row.

\subsection{Oracle generation (evaluation only)}
\label{sec:method:oracle}

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{lcc}
    \toprule
    \textbf{Setting} & \textbf{Oracle (labels)} & \textbf{Prefill probe (signals)} \\
    \midrule
    Decoding & Greedy ($\tau=0$) & No generation \\
    Distribution & --- & $\tau=1$; full-vocab softmax \\
    Purpose & Offline $y(q,M_i)$, $y^{\text{opp}}$ & $c(q)$, $H$, $m$ \\
    \bottomrule
  \end{tabular}
  \caption{Oracle full inference vs.\ prefill probe.}
  \label{tab:oracle-vs-probe}
\end{table}

\noindent
Signals and oracle labels are strictly separated: no information from $y$ enters signal computation.

\subsection{Scope}
Locked pool, datasets, split policy, calibration procedure for $c(q)$, sample sizes, and evaluation metrics are in \S\ref{sec:setup}.
The routing policy (\S\ref{sec:routing}) is defined only after \S\ref{sec:characterization} establishes what routing-relevant information the signals carry.
