% §7 Discussion

\subsection{Synthesis}
\label{sec:disc:synthesis}

We characterized pre-hoc unsupervised routing signals for multi-LLM selection: what information the model-independent and model-dependent signal families carry, whether they provide complementary predictive information (Study III), and whether that prediction supports lightweight routing (Study IV).
Our scientific contribution is \textbf{characterization and prediction from lightweight signals}---not a learned pre-hoc estimator \citep{cao2026scope} or deployable routing product \citep{ong2024routellm,guha2024smoothie}.

\textbf{[TBD: per-family informativeness; $\Delta$AUROC; routing utility across $\lambda$.]}

\subsection{Prediction then decision}
\label{sec:disc:pipeline}

Separating \textbf{prediction} (Studies I--III) from \textbf{decision} (Study IV) clarifies what the signals afford.
Study III asks whether families jointly estimate routing opportunity; Study IV asks whether that estimate improves assignment under controllable cost preferences.
Negative prediction does not invalidate the paper; it bounds what unsupervised pre-hoc probes can support.

\subsection{Where signals fail}
\label{sec:disc:failures}

\textbf{[TBD: error analysis by oracle bucket.]}
We distinguish false escalation, missed opportunity, and queries where routing is unnecessary (easy) or unhelpful (too-hard).
Failures may reflect insufficient weak--strong tier gap, not only weak probes \citep{hu2024routerbench}.

\subsection{Timing and probe cost}
\label{sec:disc:timing}

Post-generation unsupervised routers \citep{guha2024smoothie,niu2026cascal,mu2025ragrouting} and learned pre-hoc estimators \citep{cao2026scope} operate with different information and training budgets than pre-hoc logprob statistics (\S\ref{sec:char:cost}).
\textbf{[TBD: measured timings.]}

\subsection{Informativeness vs.\ calibration}
\label{sec:disc:calibration}

Instruction-tuned models are often miscalibrated on MCQ prompts \citep{he2023uncertainty,plaut2024chat}, yet logits may still rank routing need.
We report $\rho$ and AUROC as primary metrics; deployment would require separate calibration \citep{kotte2026ucci}.

\subsection{Limitations}
\label{sec:disc:limits}

\begin{itemize}
  \item Fixed binary pool in one model family.
  \item First-token prefill probes only; broader $\mathcal{S}$ deferred.
  \item Offline oracle labels for evaluation; no human preference annotations.
  \item MMLU subjects chosen by feasibility \citep{hu2024routerbench}.
\end{itemize}

\subsection{Future work}
\label{sec:disc:future}

Additional signal families; larger pools; conformal routing \citep{su2025cprouter}; RouterBench-scale comparisons \citep{hu2024routerbench}; contrast with supervised activation probes \citep{lugoloobi2026failures}.
Retrieval-augmented calibration and learned pre-hoc estimators \citep{cao2026scope} are orthogonal system directions, not extensions of this characterization study.
